Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2025-02-01 paper.qmd
spell checked

This paper describes a new projection pursuit index to find projections where a new sample is most distant from the existing normal distribution, that is integrated with the guided tour to visualise the ways that the data departs from the reference distribution. This work also expands on current outlier detection methods such as @Kandanaarachchi02012021 and @loperfido by providing integrated multivariate visualisation to examine the outlyingness in different projections. It also provides the ability to examine potential outliers interactively - when they are outlying in different ways it can interfere with detection numerically, but visually one might group different types of outlyingness, and operate the outlier detection on subsets.
<!--\textcolor{red}{This is a new development, since no existing method can compare to a reference distribution. However, the new approach can also be applied to outlier detection when we estimate the reference normal distribution from the data using robust measures. Classical methods to detect outliers with projection pursuit are based on skewness (Loperfido) (even the classical indexes were supposed to detect outliers I think?).... also dobin package by Sevvandi might be relevant here. XXX when/why might it be better to use our method?}-->

@@

Let $P$ be a ($p\times 2$) matrix, whose columns form an orthonormal basis for the \gD{} projection space. Let $\mathbfit{y}$ be a \gD{} vector, and let $\mathbfit{\mu}_p := \mathbfit{\mu} P$ denote the projected mean.
\begin{theorem}
  The projection of the \pD{} ellipsoid decribed in Equation~\ref{eq:pd} onto the \gD{} projection space described by $P$ has the equation
\begin{equation}
(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.
\end{equation}
\end{theorem}
\begin{proof}
The projection of a hyperellipsoid onto \gD{} space is an ellipse. The boundary of this \gD{} ellipse is the orthogonal projection of select points on the boundary of the hyperellipsoid. For $\mathbfit{x}$ on the hyperellipsoid, its orthogonal projection $\mathbfit{x}P$ is on the boundary of the projected ellipse if and only if the tangent plane to the hyperellipsoid at $\mathbfit{x}$ is perpendicular to the projection plane described by $P$. As the hyperellipsoid is a level curve, described by Equation~(\ref{eq:pd}), this means the gradient of the left hand side of Equation~(\ref{eq:pd}) at $\mathbfit{x}$ is parallel to the projection plane. Note that any vector parallel to the projection plane can be expressed as $2\mathbfit{s}P^T$ for some $\mathbfit{s} \in \mathbb{R}^2$.

@@

We denote points on the boundary of the projected ellipse by $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for some $\mathbfit{x}$ satifying Equation~(\ref{eq:grad}). Then, as $\mathbfit{\mu}_p = \mathbfit{\mu} P$ by definition, $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, again substituting $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$. We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in Equation~(\ref{eq:c2}). From this we can compute the analog equation for the projection as

@@

This example is motivated by a problem posed during consulting with a pharmaceutical company. The data shown here is simulated, for privacy purposes, but reflects the patterns seen in the clinical data. Liver function tests commonly provide measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase (GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT). There are normal ranges on these measurements reported by @lib-med-liver-norms and listed in @tbl-liver-norms.

@@

For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum teperature (atmn), while there is disagreement between outlyingness in other variables. Note that compared to the years 1965 and 1966 they have unusually low minimum temperature (atmn) while having average values for the number of days with maximum temperature above 25°C (nsm), and this contrast is also captured in the projection.

@@

Reproduction of the results from @MAYRHOFER2023, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm [@MAYRHOFER2023]. Red cells indicate values higher than expected, blue cells values lower than expected. For years that have been identified as outliers in our approach we show the cluster assignment in the last row.
:::

@@

This paper has provided a new projection pursuit index for comparing a sample of data against a multivariate normal reference distribution. It has also provided an analytical result for drawing \gD{} projections of the \pD{} ellipse corrsponding to the normal reference distribution. These combined provide new ways to visualise multivariate data, for this particular scenario.

>>>>

This paper describes a new projection pursuit index to find projections where a new sample is most distant from the existing normal distribution, that is integrated with the guided tour to visualise the ways that the data departs from the reference distribution. This work also expands on current outlier detection methods such as @Kandanaarachchi02012021 and @loperfido by providing integrated multivariate visualisation to examine the outlyingness in different projections. It also provides the ability to examine potential outliers interactively - when they are outlying in different ways it can interfere with detection numerically, but visually one might group different types of outlyingness, and operate the outlier detection on subsets.

@@

Let $P$ be a ($p\times 2$) matrix, whose columns form an orthonormal basis for the \gD{} projection space. Let $\mathbfit{y}$ be a \gD{} vector, and let $\mathbfit{\mu}_p := \mathbfit{\mu} P$ denote the projected mean.
\begin{theorem}
  The projection of the \pD{} ellipsoid described in Equation~\ref{eq:pd} onto the \gD{} projection space described by $P$ has the equation
\begin{equation}
(\mathbfit{y} - \mathbfit{\mu}_p)(P^T \Sigma P)^{-1}(\mathbfit{y} - \mathbfit{\mu}_p)^T = c^2.
\end{equation}
\end{theorem}
\begin{proof}
The projection of a hyperellipsoid onto \gD{} space is an ellipse. The boundary of this \gD{} ellipse is the orthogonal projection of select points on the boundary of the hyperellipsoid. For $\mathbfit{x}$ on the hyperellipsoid, its orthogonal projection $\mathbfit{x}P$ is on the boundary of the projected ellipse if and only if the tangent plane to the hyperellipsoid at $\mathbfit{x}$ is perpendicular to the projection plane described by $P$. As the hyperellipsoid is a level curve, described by Equation~(\ref{eq:pd}), this means the gradient of the left hand side of Equation~(\ref{eq:pd}) at $\mathbfit{x}$ is parallel to the projection plane. Note that any vector parallel to the projection plane can be expressed as $2\mathbfit{s}P^T$ for some $\mathbfit{s} \in \mathbb{R}^2$.

@@

We denote points on the boundary of the projected ellipse by $\mathbfit{y}$, so that $\mathbfit{y} = \mathbfit{x}P$ for some $\mathbfit{x}$ satisfying Equation~(\ref{eq:grad}). Then, as $\mathbfit{\mu}_p = \mathbfit{\mu} P$ by definition, $\mathbfit{y} - \mathbfit{\mu}_p = (\mathbfit{x} - \mathbfit{\mu})P = \mathbfit{s} P^T \Sigma P$, again substituting $\mathbfit{x}-\mathbfit{\mu} = \mathbfit{s} P^T \Sigma$. We then substitute $(\mathbfit{y} - \mathbfit{\mu}_p) (P^T \Sigma P)^{-1}$ for $\mathbfit{s}$ in Equation~(\ref{eq:c2}). From this we can compute the analog equation for the projection as

@@

This example is motivated by a problem posed during consulting with a pharmaceutical company. The data shown here is simulated, for privacy purposes, but reflects the patterns seen in the clinical data. Liver function tests commonly provide measurements on albumin, protein, bilirubin, gamma-glutamyl-transferase (GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and alanine aminotransferase (ALT). There are normal ranges on these measurements reported by @lib-med-liver-norms and listed in @tbl-liver-norms.

@@

For example we see that cluster 4 contains the years 1962 and 1996, and for these two years the result from @MAYRHOFER2023 also indicates unusually low values for the minimum temperature (atmn), while there is disagreement between outlyingness in other variables. Note that compared to the years 1965 and 1966 they have unusually low minimum temperature (atmn) while having average values for the number of days with maximum temperature above 25°C (nsm), and this contrast is also captured in the projection.

@@

Reproduction of the results from @MAYRHOFER2023, showing which cells have been identified as outlying by their Shapley Cell Detector algorithm [@MAYRHOFER2023]. Red cells indicate values higher than expected, blue cells values lower than expected. For years that have been identified as outliers in our approach we show the cluster assignment in the bottom row.
:::

@@

This paper has provided a new projection pursuit index for comparing a sample of data against a multivariate normal reference distribution. It has also provided an analytical result for drawing \gD{} projections of the \pD{} ellipse corresponding to the normal reference distribution. These combined provide new ways to visualise multivariate data, for this particular scenario.
